---
title: "Lab 09: Algorithmic Bias"
author: "Thomas Huang"
date: "2026-03-26"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("/Users/roberthuang/Desktop/Wake/Spring 2026/DSPsych/lab-09-ethics-algorithmic-bias/data/compas-scores-2-years.csv")
```

# Part 1 

```{r}
# Load the COMPAS data
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

## Ex 1

```{r}
nrow(compas) #There are 7,214 rows in the dataset, each of which represents a convinction.
ncol(compas) #There are 53 columns in the dataset.
```

# Ex 2 

```{r}
n_distinct(compas$id) #By looking at the unique ids, there are 7,214 unique defendants in the dataset, which is the same as the number of rows. If not, a reason might be an individual having multiple convictions on the record.
```

# Ex 3 

The distribution of risk scores is right skewed.
```{r}
ggplot(compas, aes(x = decile_score)) +
  geom_histogram(binwidth = 1)
```

## Ex 4

```{r}
ggplot(compas, aes(x = race, fill = sex)) +
  geom_bar() +
  facet_wrap(~age_cat, ncol = 2) +
  coord_flip()
```

# Part 2
## Ex 5

There seems to be a small difference where higher scores predict actual recidivism. But more information is needed for inference.
```{r}
compas <- compas %>% 
  mutate(two_year_recid = as.factor(two_year_recid))

ggplot(compas, aes(x = two_year_recid, y = decile_score)) +
  geom_col(binwidth = 1)
```

## Ex 6

```{r}
compas <- compas %>% 
  mutate(compas_classification =
  case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 4 & two_year_recid == 1 ~ "FN",
    decile_score <= 4 & two_year_recid == 0 ~ "TN"
))
```

## Ex 7

68.4% of its predicitons are correct. This seems to be an okay performance, above the by-change rate but not impressively high.,
```{r}
prop.table(table(compas$compas_classification))
```




