---
title: "Lab 09: Algorithmic Bias"
author: "Thomas Huang"
date: "2026-03-26"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
library(gmodels)
library(lmtest)
library(caret)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("/Users/roberthuang/Desktop/Wake/Spring 2026/DSPsych/lab-09-ethics-algorithmic-bias/data/compas-scores-2-years.csv")
```

# Part 1 

```{r}
# Load the COMPAS data
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

## Ex 1

```{r}
nrow(compas) #There are 7,214 rows in the dataset, each of which represents a convinction.
ncol(compas) #There are 53 columns in the dataset.
```

# Ex 2 

```{r}
n_distinct(compas$id) #By looking at the unique ids, there are 7,214 unique defendants in the dataset, which is the same as the number of rows. If not, a reason might be an individual having multiple convictions on the record.
```

# Ex 3 

The distribution of risk scores is right skewed.
```{r}
ggplot(compas, aes(x = decile_score)) +
  geom_histogram(binwidth = 1)
```

## Ex 4

```{r}
ggplot(compas, aes(x = race, fill = sex)) +
  geom_bar() +
  facet_wrap(~age_cat, ncol = 2) +
  coord_flip()
```

# Part 2
## Ex 5

There seems to be a small difference where higher scores predict actual recidivism. But more information is needed for inference.
```{r}
compas <- compas %>% 
  mutate(two_year_recid = as.factor(two_year_recid))

ggplot(compas, aes(x = two_year_recid, y = decile_score)) +
  geom_col(binwidth = 1)
```

## Ex 6

```{r}
compas <- compas %>% 
  mutate(compas_classification =
  case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 4 & two_year_recid == 1 ~ "FN",
    decile_score <= 4 & two_year_recid == 0 ~ "TN"
))
```

## Ex 7

68.4% of its predicitons are correct. This seems to be an okay performance, above the by-change rate but not impressively high.,
```{r}
prop.table(table(compas$compas_classification))
```

# Part 3
## Ex 8

The distribution of decile scores among white defendants is right skewed, which means white defendants are lessly likely to get a higher decile score. But black defendants have a similar chance of getting lower or higher decile scores.
```{r}
compas_r <- compas %>% 
  filter(race %in% c("African-American", "Caucasian"))

ggplot(compas_r, aes(x = decile_score)) +
  geom_histogram(binwidth = 1) +
  facet_wrap(~ race)
```

## Ex 9

A higher percentage of black defendants are classified as high risk than that of white defendants.
```{r}
compas_r <- compas_r%>% 
  mutate(high_risk =
  case_when(
    decile_score >= 7 ~ "TRUE",
    decile_score <= 4 ~ "FALSE"
))

CrossTable(table(compas_r$high_risk, compas_r$race),expected=T,prop.r=F,prop.c=T,prop.t=F,prop.chisq=F,chisq=T,fisher=T,format="SPSS")
```

## Ex 10

False positive rate of black defendants: 31.11%.
False positive rate of white defendants: 10.67%.

False negative rate of black defendants: 35.23%.
False negative rate of white defendants: 61.96%.
```{r}
# False positive
non_recidivists <- compas_r %>%
  filter(two_year_recid == 0)

CrossTable(table(non_recidivists$high_risk, non_recidivists$race),expected=T,prop.r=F,prop.c=T,prop.t=F,prop.chisq=F,chisq=T,fisher=T,format="SPSS")

# False negative
recidivists <- compas_r %>%
  filter(two_year_recid == 1)

CrossTable(table(recidivists$high_risk, recidivists$race),expected=T,prop.r=F,prop.c=T,prop.t=F,prop.chisq=F,chisq=T,fisher=T,format="SPSS")
```

## Ex 11

The overall accuracy among black defendants is lower. There's a higher false positive rate among black defendants and a higher false negative rate among white defendants.
```{r}
compas_r %>% 
  filter(!is.na(compas_classification)) %>% 
  ggplot(aes(x = race, fill = compas_classification)) +
  geom_bar(position = "fill") +
  labs(
    fill = "Prediction Type",
    y = "Proportion"
  )+
  scale_fill_discrete(
  labels = c("TP" = "True Positive",
             "FP" = "False Positive",
             "FN" = "False Negative",
             "TN" = "True Negative")
)
```

# Part 4
## Ex 12

The algorithm assigns a much heavier weight on prior convictions among black defendants.
```{r}
ggplot(compas_r, aes(x = priors_count, y = decile_score, color = race)) +
  geom_point()
```

## Ex 13

I first regressed the two-year recidivation on the risk score (model 1), and then added race (model 2) and the interaction term of race and the risk score (model 3). Across three models, the risk score is a significant predictor of the two-year recidivation. Race and the interaction term are not significant predictors of the outcome. Results of likelihood ratio tests indicate the three models have similar model fit. Therefore, the risk score predicts two-year recidivation in a similar way among black and white defendants. In other words, there is not enough evidence to show the relationship between the risk score and two-year recidivation is related to the race of defendants. From this perspective, Northpointe’s claim that the algorithm is fair seems to be supported by the results
```{r}
model1 <- glm(two_year_recid ~ decile_score, data = compas_r, family = binomial)
summary(model1)

model2 <- glm(two_year_recid ~ decile_score + race, data = compas_r, family = binomial)
summary(model2)

model3 <- glm(two_year_recid ~ decile_score + race + decile_score*race, data = compas_r, family = binomial)
summary(model3)

lrtest(model1, model2)
lrtest(model2, model3)
lrtest(model1, model3)
```

# Part 5
## Ex 14
I would ensure equal or at least similar distribution of TN, TP, FN, and FP rates across races.

## Ex 15
Ensuring similar prediction accuracy across races may compromise calibration and vice versa.

## Ex 16
Algorithm developers should give a full disclosure of the caveats when using such information. In the case of COMPAS, Northpointe should disclose their model has different prediction accuracies across races.

# Stretch
## Ex 17

Perhaps the distribution of prior convictions among black defendants are more right skewed. So the algorithm is more likely to assume black defendants will recidivate.
```{r}
ggplot(compas, aes(x = priors_count, fill = race)) +
  geom_histogram(binwidth = 1)
```

## Ex 18

The visualizations show that black defendants were convicted at younger ages with felony charges, which may contribute to the biases in the algorithm.
```{r}
ggplot(compas, aes(x = age, fill = race)) +
  geom_histogram(binwidth = 1)

ggplot(compas, aes(x = c_charge_degree, fill = race)) +
  geom_bar(binwidth = 1)
```

## Ex 19

```{r}
# Create a logistic regression model
recid_model <- glm(
  two_year_recid ~ age + priors_count + c_charge_degree,
  data = compas,
  family = binomial()
)

# Add predicted probabilities to the dataset
compas <- compas %>%
  mutate(
    predicted_prob = predict(recid_model, type = "response"),
    our_high_risk = predicted_prob >= 0.5
  )
```

## Ex 20

The model is still kind of biased. The false positive rate is still much higher among black defendants by looking at the specificity.
```{r}
compas$two_year_recid_bool <- compas$two_year_recid == 1
compas$two_year_recid_bool <- as.factor(compas$two_year_recid_bool)
compas$our_high_risk <- as.factor(compas$our_high_risk)

# Black defendants
compas_black_defendants <- compas %>% 
  filter(race == "African-American")

set.seed(304844)
confusionMatrix(
  compas_black_defendants$our_high_risk,
  compas_black_defendants$two_year_recid_bool,
  positive = NULL,
  prevalence = NULL,
  mode = "sens_spec",
)

# White defendants
compas_white_defendants <- compas %>% 
  filter(race == "Caucasian")

set.seed(304844)
confusionMatrix(
  compas_white_defendants$our_high_risk,
  compas_white_defendants$two_year_recid_bool,
  positive = NULL,
  prevalence = NULL,
  mode = "sens_spec",
)
```

## Ex 21

```{r}
# Create a logistic regression model with race
recid_model_with_race <- glm(
  two_year_recid ~ age + priors_count + c_charge_degree + race,
  data = compas,
  family = binomial()
)

# Add predicted probabilities to the dataset
compas <- compas %>%
  mutate(
    predicted_prob_with_race = predict(recid_model_with_race, type = "response"),
    race_high_risk = predicted_prob_with_race >= 0.5
  )
```

## Ex 22

The model is still kind of biased. Oddly, the predictions of white defendants are exactly the same across the two models. See the last confusion matrix.
```{r}
compas$race_high_risk <- as.factor(compas$race_high_risk)

# Black defendants
compas_black_defendants <- compas %>% 
  filter(race == "African-American")

set.seed(304844)
confusionMatrix(
  compas_black_defendants$race_high_risk,
  compas_black_defendants$two_year_recid_bool,
  positive = NULL,
  prevalence = NULL,
  mode = "sens_spec",
)

# White defendants
compas_white_defendants <- compas %>% 
  filter(race == "Caucasian")

set.seed(304844)
confusionMatrix(
  compas_white_defendants$race_high_risk,
  compas_white_defendants$two_year_recid_bool,
  positive = NULL,
  prevalence = NULL,
  mode = "sens_spec",
)

confusionMatrix(
  compas_white_defendants$race_high_risk,
  compas_white_defendants$our_high_risk,
  positive = NULL,
  prevalence = NULL,
  mode = "sens_spec",
)
```

## Ex 23

Given what we have here, such algorithms should not be used. I think the biggest concern is accuracy, which is in the 60% to 70% range. This is far from being perfect. If they are used, decision-makers should be told the key performance indices, such as accuracy, sensitivity, specificity, and fairness. The trade-off between accuracy and fairness seems to be tricky to handle. One potential solution is a multilevel model structure, such as including racial-level and defendant-level predictors. The role of transparency in algorithmic decision-making is to acknowledge there is no transparency in algorithmic decision-making.
