---
title: "Lab 09: Algorithmic Bias"
author: "Thomas Huang"
date: "2026-03-26"
output: github_document
---

## Load Packages and Data  

First, let's load the necessary packages:  

```{r load-packages, message = FALSE}
library(tidyverse)
library(fairness)
library(janitor)
library(gmodels)
```

### The data

For this lab, we'll use the COMPAS dataset compiled by ProPublica. The data has been preprocessed and cleaned for you. You'll have to load it yourself. The dataset is available in the `data` folder, but I've changed the file name from `compas-scores-two-years.csv` to `compas-scores-2-years.csv`. I've done this help you practice debugging code when you encounter an error.

```{r}
compas <- read_csv("/Users/roberthuang/Desktop/Wake/Spring 2026/DSPsych/lab-09-ethics-algorithmic-bias/data/compas-scores-2-years.csv")
```

# Part 1 

```{r}
# Load the COMPAS data
compas <- read_csv("data/compas-scores-2-years.csv") %>%
  clean_names() %>%
  rename(
    decile_score = decile_score_12,
    priors_count = priors_count_15
  )

# Take a look at the data
glimpse(compas)
```

## Ex 1

```{r}
nrow(compas) #There are 7,214 rows in the dataset, each of which represents a convinction.
ncol(compas) #There are 53 columns in the dataset.
```

# Ex 2 

```{r}
n_distinct(compas$id) #By looking at the unique ids, there are 7,214 unique defendants in the dataset, which is the same as the number of rows. If not, a reason might be an individual having multiple convictions on the record.
```

# Ex 3 

The distribution of risk scores is right skewed.
```{r}
ggplot(compas, aes(x = decile_score)) +
  geom_histogram(binwidth = 1)
```

## Ex 4

```{r}
ggplot(compas, aes(x = race, fill = sex)) +
  geom_bar() +
  facet_wrap(~age_cat, ncol = 2) +
  coord_flip()
```

# Part 2
## Ex 5

There seems to be a small difference where higher scores predict actual recidivism. But more information is needed for inference.
```{r}
compas <- compas %>% 
  mutate(two_year_recid = as.factor(two_year_recid))

ggplot(compas, aes(x = two_year_recid, y = decile_score)) +
  geom_col(binwidth = 1)
```

## Ex 6

```{r}
compas <- compas %>% 
  mutate(compas_classification =
  case_when(
    decile_score >= 7 & two_year_recid == 1 ~ "TP",
    decile_score >= 7 & two_year_recid == 0 ~ "FP",
    decile_score <= 4 & two_year_recid == 1 ~ "FN",
    decile_score <= 4 & two_year_recid == 0 ~ "TN"
))
```

## Ex 7

68.4% of its predicitons are correct. This seems to be an okay performance, above the by-change rate but not impressively high.,
```{r}
prop.table(table(compas$compas_classification))
```

# Part 3
## Ex 8

The distribution of decile scores among white defendants is right skewed, which means white defendants are lessly likely to get a higher decile score. But black defendants have a similar chance of getting lower or higher decile scores.
```{r}
compas_r <- compas %>% 
  filter(race %in% c("African-American", "Caucasian"))

ggplot(compas_r, aes(x = decile_score, fill = race)) +
  geom_histogram(binwidth = 1)
```

## Ex 9

A higher percentage of black defendants are classified as high risk than that of white defendants.
```{r}
compas_r <- compas_r%>% 
  mutate(high_risk =
  case_when(
    decile_score >= 7 ~ "TRUE",
    decile_score <= 4 ~ "FALSE"
))

CrossTable(table(compas_r$high_risk, compas_r$race),expected=T,prop.r=F,prop.c=T,prop.t=F,prop.chisq=F,chisq=T,fisher=T,format="SPSS")
```

## Ex 10

False positive rate of black defendants: 31.11%.
False positive rate of white defendants: 10.67%.

False negative rate of black defendants: 35.23%.
False negative rate of white defendants: 61.96%.
```{r}
# False positive
non_recidivists <- compas_r %>%
  filter(two_year_recid == 0)

CrossTable(table(non_recidivists$high_risk, non_recidivists$race),expected=T,prop.r=F,prop.c=T,prop.t=F,prop.chisq=F,chisq=T,fisher=T,format="SPSS")

# False negative
recidivists <- compas_r %>%
  filter(two_year_recid == 1)

CrossTable(table(recidivists$high_risk, recidivists$race),expected=T,prop.r=F,prop.c=T,prop.t=F,prop.chisq=F,chisq=T,fisher=T,format="SPSS")
```

## Ex 11

The overall accuracy among black defendants is lower. There's a higher false positive rate among black defendants and a higher false negative rate among white defendants.
```{r}
compas_r %>% 
  filter(!is.na(compas_classification)) %>% 
  ggplot(aes(x = race, fill = compas_classification)) +
  geom_bar(position = "fill") +
  labs(
    fill = "Prediction Type",
    y = "Proportion"
  )+
  scale_fill_discrete(
  labels = c("TP" = "True Positive",
             "FP" = "False Positive",
             "FN" = "False Negative",
             "TN" = "True Negative")
)
```

# Part 4